---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(knitr)
library(dplyr)
library(tibble)
```

# icd10es - A user-friendly R package for disease clasification in Spanish 🇲🇽

---

## Table of Contents

- [Package description](#description)
- [Installing icd10es](#installing)
- [Features at a glance](#features-at-a-glance)
  - [Printing information of a CIE-10 entry](#printing-info)
  - [Looking up a string in the catalog](#icdlookup)
  - [Using an external catalog](#external)
  - [Looking up entries within death certificates](#death-certificates)
- [License](#license)
- [Credits](#credits)
- [Thanks](#thanks)

## Package description 😷

`icd10es` is an R package created for Spanish-speaking Bioinformatics specialists 👩‍⚕️ who have to deal with classifying written descriptions of diseases, symptoms, and injuries, among other health-related issues, in the 10th edition of the International Statistical Classification of Diseases and Related Health Problems (ICD-10 for short), referred to as CIE-10 in Spanish. ⚕️

## Installing icd10es 

```{r, eval =FALSE, message=FALSE, warning =FALSE}
devtools::install_github("mcarmonabaez/icd10es")
```

*Congratulations!* Now you can use this package! 🎉

## Features at a glance

### Printing information of a CIE-10 entry


Let's start with a simple task: say you wish to know what the entry 'A00.0' in the catalog contains.
The function `printInfo` can help with that. Changing the value of the parameter `tabular` you
can decide whether you want to 

- get only the canonical term in table form, 

- get the canonical and all inclusion terms (if they exist) also in table form, 

- print in all associated information in the console for quick inquiries.

```{r}
library(icd10es)
printInfo('S72.1', tabular = 'single')
printInfo('S72.1', tabular = 'simple')
printInfo('S72.1', tabular = 'full')
```




### Looking up a string in the catalog 🔎

The main function of `icd10es` consists in entering a string which is expected to match some entry in the CIE-10 and finding said entry, all via the `ICDLookUp` function. The string does not have to be identical to the entry: herein lies the usefulness of the package.

The function first tries to find an __exact__ match in the catalog, but often it occurs that the string either has a typo of some kind (e.g. writing 'pnuemonia' instead of 'pneumonia) or uses a more colloquial way of referring to the disease or symptom and is not its 'full name'. When this happens, the function tries fuzzy matching using the Jaro-Winkler similarity metric. 

For example, in the CIE-10, all cancers are referred to in a more formal way, such as 'tumor maligno del colon' instead of 'cancer de colon' (in English: 'malignant neoplasm of colon' instead of 'colon cancer'). `ICDLookUp` would give the following output:

```{r}
ICDLookUp('cancer de colon', tabular = 'simple')
```
Note how the `tabular` parameter is inherited to `printInfo`.

When doing fuzzy matching, one can be more or less strict. 
This is reflected in the `jwBound` parameter of `ICDLookUp`: the Jaro-Winkler
similarity goes from 0 (no similarity) to 1 (exact match), and the default value of
`jwBound` is 0.9. That is, only entries with a similarity to the entered string equal
or higher than 0.9 will be considered. But if one finds that the function didn't find a
result, one can try lowering the bound:

```{r}
ICDLookUp('sindrome dandie-waker', jwBound = 0.9, tabular = 'simple')
ICDLookUp('sindrome dandie-waker', jwBound = 0.8, tabular = 'simple')
```


### Using an external catalog

It can happen that the user wants to look up strings in a different,
specialized catalog. This could be for example when using an auxiliary catalog
which has alternative names of some diseases due to regional variations
(like when a country or a country's province historically calls a disease in a special way).

This can be done by making the `ICDLookUp` parameter `useExternal = TRUE`,
and by giving a `dataframe` to `externalCatalog`:

```{r}
auxCatalog <- read.delim('https://raw.githubusercontent.com/mcarmonabaez/icd10es/master/inst/extdata/inputs/diabetes_subcategories.csv')
ICDLookUp('Diabetes tipo i con coma', tabular = 'simple',
          useExternal = TRUE, externalCatalog = auxCatalog)
```

### Looking up entries within death certificates 🧪

It is very common to be in possession of longer texts that describe a series of
diseases and symptoms which could be matched to the CIE-10. Some examples
include death certificates or medical records. There, a physician may list some or all
comorbidities a person presents when having a medical checkup or when passing away.
One may then wish to match all listed health-related problems with the CIE-10.

```{r}
exampleCerificates <-
  tibble::tribble(~id, ~cause,
                  1, 'HEMORRAGIA SUBARACNOIDEA. HIPERTENSION ARTERIAL SISTEMICA. DISLIPIDEMIA.',
                  2, 'INFARTO CEREBRAL, HIPERTENSION ARTERIAL SISTEMICA, TRIGLICERIDEMIA.',
                  3, 'HERIDA PRODUCIDA POR PROYECTIL DE ARMA DE FUEGO PENETRANTE DE TORAX.',
                  4, 'CHOQUE HIPOVOLEMICO, DIARREA CRONICA, INFECCION POR VIRUS DE INMUNODEFICIENCIA HUMANA.',
                  5, 'EVENTO VASCULAR CEREBRAL, ENFERMEDAD RENAL TERMINAL, DIABETES MELLITUS TIPO 2.',
                  6, 'ANEURISMA CEREBRAL, ENCEFALOPATIA HEPATICA.',
                  7, 'INFARTO AGUDO AL MIOCARDIO, CARDIOPATIA HIPERTENSIVA, HIPERTENSION ARTERIAL SISTEMICA',
                  8, 'MENINGIOMA, HIPERTENSION ARTERIAL SISTEMICA.',
                  9, 'ENCEFALOPATIA HEPATICA, CIRROSIS HEPATICA, ALCOHOLISMO CRONICO',
                  10, 'INFARTO AGUDO AL MIOCARDIO, DIABETES MELLITUS TIPO II.'
  )
exampleCerificates
```
First, one would have to _tokenize_ each entry in the certificate, creating a long
`dataframe` in the following way using `tokenizeCertificates`:

```{r}
tokenizedCerificates <- tokenizeCertificates(exampleCerificates)
print(tokenizedCerificates, n = Inf)
```

One can then proceed to use `ICDLookUp` to try to find an entry in the catalog 
for each of the entries in the certificate:
```{r message=FALSE, warning=FALSE, results='hide'}
results <- lapply(unique(tokenizedCerificates$id),
                     function(x) {
                       print(x)
                       subset <- dplyr::filter(tokenizedCerificates, id == x)
                       lapply(subset$cause, ICDLookUp)
                     }) %>%
  bind_rows(.id = 'id',) %>%
  mutate(id = as.numeric(id)) %>%
  arrange(id) %>%
  group_by(id) %>%
  mutate(order = row_number()) %>%
  ungroup() 
```


```{r}
tokenizedCerificates$result <- results$disease
print(tokenizedCerificates, n = Inf)
```

## License

This package is made available under the MIT License.

## Credits

This package is created and maintained by Mariana Carmona-Baez and Juan Bernardo Martínez Parente-Castañeda. 🐞

*We're open to suggestions, feel free to message us on mcarmonabaez@gmail.com and jbmpc@outlook.com.*
*Pull requests are also welcome!* 🔀


## Thanks

Thanks to Christopher Ormsby for his input and for letting us be part of this awesome project :hospital:

Thanks to [Teresa Ortiz](https://github.com/tereom) for her invaluable guidance :crystal_ball:




